---
title: "City Proximity and Land Cover Analysis for California Wildfires (2020)"
output: html_document
date: "2025-03-12"
---

# Motivation

The 2020 California wildfire season was one of the most destructive on record, with over 4 million acres burned across the state. This part of analysis focuses on two critical determinants of wildfire impact:

(1) the proximity of fires to populated areas.

(2) the types of land cover affected by these fires. These factors are essential for understanding both the human impact and ecological significance of wildfire events.

# City Proximity Analysis

Understanding the spatial relationship between wildfire perimeters and population centers is crucial for several reasons:

-   Emergency Response Planning: By quantifying the distance between fires and communities, emergency management can better prioritize resources and evacuation efforts for the most at-risk populations.

-   Public Health Implications: Even communities not directly threatened by flames can experience significant health impacts from smoke and air quality degradation, which are influenced by proximity to active fires.

-   Urban-Wildland Interface: As California's development continues to expand into previously wild areas, the interface between human settlements and fire-prone landscapes becomes increasingly important to map and understand.

Our analysis uses geospatial techniques to measure distances between city centroids and fire perimeters, identifying communities within critical distance thresholds (particularly those within 50km of fire boundaries).

# Data Sources

-   California Fire Perimeters Dataset. accessed at: <https://data.cnra.ca.gov/dataset/california-fire-perimeters-all?utm_source=chatgpt.com>

-   Administrative level 1 boundaries, which include states, provinces, and regions within countries from Natural Earth (scale 1:110)

-   TIGER/Line Shapefiles, which are provided by the U.S. Census Bureau containing cities and towns in LA. accessed at: <https://www.census.gov/cgi-bin/geo/shapefiles/index.php?year=2020&layergroup=Places>

-   Annuals NLCD Land Cover Data. Accessed at: <https://www.mrlc.gov/data?f%5B0%5D=category%3ALand%20Cover&f%5B1%5D=project_tax_term_term_parents_tax_term_name%3AAnnual%20NLCD>

```{r, message=FALSE, warning=FALSE}
#Distance from cities
##########

# Load necessary libraries
# install.packages(c("sf", "terra", "raster", "httr", "jsonlite","FedData"))
library(sf)
library(dplyr)
library(ggplot2)
library(matrixStats)
library(terra)
library(raster)
library(httr)
library(jsonlite)
library(FedData)
```

```{r, message=FALSE}
# Read Fire Data (2020 only)
data <- st_read("~/Desktop/Victoria/DSDM/Term2/Geospatial/Final_Project/California_Fire_Perimeters_(all)/California_Fire_Perimeters_(all).shp")
data <- data %>% filter(YEAR_ == 2020)

# Read California State Boundary
US_States <- st_read("~/Desktop/Victoria/DSDM/Term2/Geospatial/Final_Project/ne_110m_admin_1_states/ne_110m_admin_1_states_provinces.shp")
US_States_simple <- US_States %>% dplyr::select(name, postal, latitude, longitude, geometry)
CA <- US_States_simple %>% filter(postal == "CA")

# Read the “Places” shapefile
# Loading this to get info about cities
places <- st_read("~/Desktop/Victoria/DSDM/Term2/Geospatial/Final_Project/tl_2020_06_place/tl_2020_06_place.shp")
colnames(places)
head(places)
```

```{r, warning=FALSE}
# Reproject all layers to EPSG:3310 at the start
data_3310 <- st_transform(data, 3310)
CA_3310 <- st_transform(CA, 3310)
places_ca_centroids_3310 <- st_centroid(st_transform(places, 3310)) 

# Combine all fire polygons into one and compute centroid
fire_union_3310 <- st_union(data_3310)
fire_union_centroid_3310 <- st_centroid(fire_union_3310)
```

```{r, warning=FALSE}
# Distance from each city centroid to the fire boundary and union centroid
places_ca_centroids_3310$dist_to_fire_boundary <- st_distance(places_ca_centroids_3310, fire_union_3310)
places_ca_centroids_3310$dist_to_fire_union_centroid <- st_distance(places_ca_centroids_3310, fire_union_centroid_3310)

# Compute centroids of individual fires and their distances to cities
fire_centroids_3310 <- st_centroid(data_3310)
city_fire_matrix <- st_distance(places_ca_centroids_3310, fire_centroids_3310)

# Extract the nearest fire centroid distance per city
places_ca_centroids_3310$dist_to_nearest_fire_centroid <- rowMins(city_fire_matrix)
```

```{r}
ggplot() +
  # California boundary
  geom_sf(data = CA_3310, fill = "white", color = "black") +
  
  # Fire polygons
  geom_sf(data = data_3310, fill = "red", alpha = 0.6, color = NA) +
  
  # City centroids, colored by distance to boundary
  geom_sf(
    data = places_ca_centroids_3310,
    aes(color = as.numeric(dist_to_fire_boundary)),
    size = 1
  ) +
  
  scale_color_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    color = "Distance to Fire (m)",
    title = "California Fires 2020: Distance from Cities to Fire Perimeters (EPSG:3310)"
  )
```

```{r}
max_dist <- max(as.numeric(places_ca_centroids_3310$dist_to_fire_boundary), na.rm = TRUE)

ggplot() +
  geom_sf(data = CA_3310, fill = "white", color = "black") +
  geom_sf(data = data_3310, fill = "black", alpha = 0.9, color = NA) +
  geom_sf(
    data = places_ca_centroids_3310,
    aes(color = as.numeric(dist_to_fire_boundary)),
    size = 1
  ) +
  scale_color_gradientn(
    colors = c("#FF0000", "#FFA500", "#808080", "#008000"),
    breaks = c(0, 1000, 5000, 10000, 50000, 100000, 150000, max_dist),
    labels = c("0 km", "1 km", "5 km", "10 km", "50 km", "100 km", "150 km",
               paste0(round(max_dist/1000), " km"))
  ) +
  guides(
    # Make the colorbar bigger/taller, with some padding
    color = guide_colorbar(
      barwidth = 1.5,    # width of the bar in "plot" units
      barheight = 10,    # height of the bar in "plot" units
      title.position = "top"  # puts the legend title on top
    )
  ) +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 10)
  ) +
  labs(
    color = "Distance to Fire (m)",
    title = "CA Fires 2020: Distance from Cities"
  )
```

```{r}
#### Fire Proximity - Filter 50km

# 1. Convert distance units to numeric (removing "units" class):
places_ca_centroids_3310$dist_num <- as.numeric(places_ca_centroids_3310$dist_to_fire_boundary)

# 2. Filter for distances less than 50 km (50,000 meters):
cities_close <- places_ca_centroids_3310 %>%
  filter(dist_num < 50000)

ggplot() +
  geom_sf(data = CA_3310, fill = "white", color = "black") +
  geom_sf(data = data_3310, fill = "black", alpha = 0.9, color = NA) +
  geom_sf(
    data = cities_close,
    aes(color = as.numeric(dist_to_fire_boundary)),
    size = 1
  ) +
  scale_color_gradientn(
    colors = c("#FF0000", "#FFA500", "#808080", "#008000")
  ) +
  guides(
    # Make the colorbar bigger/taller, with some padding
    color = guide_colorbar(
      barwidth = 1.5,    # width of the bar in "plot" units
      barheight = 10,    # height of the bar in "plot" units
      title.position = "top"  # puts the legend title on top
    )
  ) +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 10)
  ) +
  labs(
    color = "Distance to Fire (m)",
    title = "CA Fires 2020: Distance from Cities - Less than 50 km"
  )
```

```{r}
##################
# List those cities for reference

# Ensure the distance column is numeric 
places_ca_centroids_3310 <- places_ca_centroids_3310 %>%
  mutate(dist_num = as.numeric(dist_to_fire_boundary))

# Filter cities that are within 50 km (i.e., less than 50,000 m)
cities_50 <- places_ca_centroids_3310 %>%
  filter(dist_num < 50000) %>%   # 50 km threshold
  arrange(dist_num)               # Rank in ascending order (closest first)

# Extract a list with city names, distances, and coordinates
cities_list <- dplyr::select(st_drop_geometry(cities_50), NAME, dist_num)

print(cities_list)
```

# Population-Weighted Risk Analysis for California Wildfires

Population-Weighted Risk Analysis for California Wildfires is essential because it provides a more accurate representation of human impact than raw geographical proximity alone. By accounting for both a community's distance from fires and its population size, this approach highlights areas where large numbers of people are endangered, not just geographical points on a map.

```{r, warning=FALSE, message=FALSE}
# This part extends our existing analysis to include population weighting

########################################################
# Extract population data or join with Census data
########################################################

# Check if the places shapefile has population data
if("POP10" %in% colnames(places) || "POPULATION" %in% colnames(places)) {
  # If population field already exists, use it
  pop_field <- ifelse("POP10" %in% colnames(places), "POP10", "POPULATION")
  
  # Join population to your existing centroids dataframe
  places_ca_centroids_3310 <- places_ca_centroids_3310 %>%
    left_join(st_drop_geometry(places) %>% 
              dplyr::select(GEOID, NAME, !!sym(pop_field)), 
              by = c("GEOID"))
  
  # Rename the population field to be consistent
  places_ca_centroids_3310 <- places_ca_centroids_3310 %>%
    rename(Population = !!sym(pop_field))
} else {
  # If no population data is available, we need to get it from Census API
  # or another source. This example uses the tidycensus package.
  
  # Install if needed
  if(!require(tidycensus)) {
    install.packages("tidycensus")
    library(tidycensus)
  }
  
  # You'll need a Census API key - get one at https://api.census.gov/data/key_signup.html
  # census_api_key("YOUR_API_KEY")
  
  # Get population data for places in California
  ca_places_pop <- get_decennial(
    geography = "place",
    variables = "P1_001N", # Total population
    state = "CA",
    year = 2020,
    geometry = FALSE
  )
  
  # Clean up the data
  ca_places_pop <- ca_places_pop %>%
    rename(Population = value) %>%
    dplyr::select(GEOID, NAME, Population)
  
  # Join to your centroids
  places_ca_centroids_3310 <- places_ca_centroids_3310 %>%
    left_join(ca_places_pop, by = c("GEOID"))
}
```

```{r, warning=FALSE, message=FALSE}
########################################################
# Calculate risk categories based on distance and population
########################################################

# Add risk categories based on distance
places_ca_centroids_3310 <- places_ca_centroids_3310 %>%
  mutate(
    # Distance in km for readability
    distance_km = dist_num / 1000,
    
    # Risk category based on distance
    risk_category = case_when(
      distance_km <= 10 ~ "Extreme",
      distance_km <= 20 ~ "High",
      distance_km <= 35 ~ "Moderate",
      distance_km <= 50 ~ "Low",
      TRUE ~ "Minimal"
    ),
    
    # Convert risk category to numeric score for calculations
    risk_score = case_when(
      risk_category == "Extreme" ~ 5,
      risk_category == "High" ~ 4,
      risk_category == "Moderate" ~ 3,
      risk_category == "Low" ~ 2,
      TRUE ~ 1
    )
  )

# Calculate population-weighted risk metrics
places_ca_centroids_3310 <- places_ca_centroids_3310 %>%
  mutate(
    # Population at risk (simple metric)
    population_at_risk = ifelse(distance_km <= 50, Population, 0),
    
    # Population risk score (weighted by both population size and risk level)
    population_risk_score = Population * risk_score,
    
    # Risk index (normalized score between 0-100)
    risk_index = population_risk_score / max(population_risk_score, na.rm = TRUE) * 100
  )

```

```{r}
########################################################
# Create a summary table of highest-risk cities
########################################################

# Check column names first to see what's available
print("Available columns in places_ca_centroids_3310:")
print(colnames(places_ca_centroids_3310))

# Check column names first to see what's available
# Now that we see the available columns, we can select the right ones
high_risk_cities <- places_ca_centroids_3310 %>%
  arrange(desc(risk_index)) %>%
  head(20) %>%
  st_drop_geometry() %>%
  dplyr::select(
    NAME.x,               # The city name is in NAME.x
    Population,
    distance_km,
    risk_category,
    population_at_risk,
    risk_index
  ) %>%
  # Rename NAME.x to City for clarity
  rename(City = NAME.x)

# Print the table
print(high_risk_cities)

# Save the table for later use
write.csv(high_risk_cities, "high_risk_cities_2020.csv", row.names = FALSE)


```

```{r}
########################################################
# Visualize population-weighted fire risk
########################################################

# Create a map showing cities with circle size proportional to population and color by risk
ggplot() +
  # California boundary
  geom_sf(data = CA_3310, fill = "white", color = "black") +
  
  # Fire polygons
  geom_sf(data = data_3310, fill = "red", alpha = 0.6, color = NA) +
  
  # City points with size based on population and color based on risk
  geom_sf(
    data = places_ca_centroids_3310 %>% filter(!is.na(Population)),
    aes(
      color = risk_category,
      size = Population
    ),
    alpha = 0.7
  ) +
  
  # Scale settings
  scale_color_manual(
    values = c("Extreme" = "#7F0000", "High" = "#FF0000", 
               "Moderate" = "#FFA500", "Low" = "#FFFF00", "Minimal" = "#00FF00"),
    name = "Fire Risk Level"
  ) +
  scale_size_continuous(
    name = "Population",
    range = c(0.5, 5),
    labels = scales::comma
  ) +
  
  # Map styling
  theme_minimal() +
  labs(
    title = "Population-Weighted Fire Risk in California (2020)",
    subtitle = "Circle size indicates population; color indicates risk level based on distance to fires",
    caption = "Data sources: Cal Fire, U.S. Census Bureau"
  )

```

```{r}
########################################################
# Create a bar chart of population at risk by category
########################################################

# Summarize population at risk by risk category
risk_summary <- places_ca_centroids_3310 %>%
  st_drop_geometry() %>%
  group_by(risk_category) %>%
  summarize(
    total_population = sum(Population, na.rm = TRUE),
    cities_count = n()
  ) %>%
  arrange(factor(risk_category, levels = c("Extreme", "High", "Moderate", "Low", "Minimal")))

# Plot the summary
ggplot(risk_summary, aes(x = factor(risk_category, 
                                    levels = c("Extreme", "High", "Moderate", "Low", "Minimal")),
                         y = total_population)) +
  geom_bar(stat = "identity", aes(fill = factor(risk_category))) +
  geom_text(aes(label = paste0(cities_count, " cities")), 
            position = position_stack(vjust = 0.5), color = "white") +
  scale_fill_manual(
    values = c("Extreme" = "#7F0000", "High" = "#FF0000", 
               "Moderate" = "#FFA500", "Low" = "#FFFF00", "Minimal" = "#00FF00"),
    name = "Fire Risk Level"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  labs(
    title = "California Population by Wildfire Risk Level (2020)",
    subtitle = "Number of cities shown in each bar",
    x = "Risk Category",
    y = "Total Population",
    caption = "Risk categories based on distance to 2020 fire perimeters"
  ) +
  theme(axis.text.x = element_text(angle = 0))
```

*Note* that the total of around 1,400 places should not be a cause for concern in our analysis. The higher-than-expected count reflects the Census Bureau's comprehensive definition of "places" in their TIGER/Line shapefiles, which extends well beyond just incorporated cities. The dataset includes:

-   incorporated municipalities (cities, towns),
-   Census Designated Places (CDPs) which are unincorporated communities with defined boundaries,
-   various smaller settlements that have distinct geographic identities recognized by the Census Bureau

While California has 482 incorporated cities, it has over 1,000 additional Census Designated Places, bringing the total close to the number we're seeing. This comprehensive inclusion actually strengthens our analysis by providing a more complete picture of population exposure to wildfire risk across all types of human settlements in California, not just major cities.

# Land Cover Type Analysis

The second component examines the relationship between land cover types and fire perimeters, which helps illuminate:

-   Fuel Availability: Different vegetation types (forests, shrublands, grasslands) have varying combustibility and fuel characteristics that influence fire behavior.

-   Development Patterns: By distinguishing between natural and developed areas affected by fires, we can better understand the human-wildfire relationship in California.

-   Ecological Impact Assessment: Identifying which ecosystem types were most affected provides insight into potential ecological consequences and recovery trajectories.

Using the National Land Cover Database (NLCD), our analysis categorizes and quantifies the types of land affected by the 2020 fires, with special attention to the distribution between developed (urban/populated) versus natural/other land covers.

```{r, message=FALSE, warning=FALSE}
#########################################
# NLCD and Fire Perimeter Data Processing
#########################################

# Load NLCD raster
nlcd_raster <- rast("/Users/macbookpro/Desktop/Victoria/DSDM/Term2/Geospatial/Final_Project/Annual_NLCD_LndCov_2020_CU_C1V0.tif")

# Load California boundary and reproject to match raster CRS
CA <- st_read("~/Desktop/Victoria/DSDM/Term2/Geospatial/Final_Project/ne_110m_admin_1_states/ne_110m_admin_1_states_provinces.shp") %>%
  filter(postal == "CA") %>%
  st_transform(crs(nlcd_raster))

# Convert California boundary to terra format and crop raster
nlcd_ca <- crop(nlcd_raster, vect(CA), mask = TRUE)
```

```{r}
# Visualize and save cropped raster
plot(nlcd_ca, main = "NLCD Land Cover - California (2020)")
```

```{r, warning=FALSE, message=FALSE}
writeRaster(nlcd_ca, "NLCD_2020_California.tif", overwrite = TRUE)

# Load California Land Cover Raster
nlcd_ca <- rast("/Users/macbookpro/NLCD_2020_California.tif")

# Load and clean 2020 fire perimeters, ensuring CRS consistency
fire_perimeters <- st_read("~/Desktop/Victoria/DSDM/Term2/Geospatial/Final_Project/California_Fire_Perimeters_(all)/California_Fire_Perimeters_(all).shp") %>%
  filter(YEAR_ == 2020) %>%
  st_make_valid() %>%
  filter(!st_is_empty(.)) %>%
  st_transform(crs(nlcd_raster))

# Convert fire perimeters to terra vector format
fire_vect <- vect(fire_perimeters)

# Extract land cover data intersecting fires
landcover_fire_extract <- terra::extract(nlcd_ca, fire_vect)

# Convert extracted data to a dataframe and rename columns
landcover_fire_data <- data.frame(landcover_fire_extract)
colnames(landcover_fire_data) <- c("ID", "Land_Cover_Type")

# Summarize extracted land cover data
landcover_summary <- landcover_fire_data %>%
  group_by(Land_Cover_Type) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

print(landcover_summary)

# # Convert raster to dataframe for plotting
# nlcd_df <- as.data.frame(nlcd_ca, xy = TRUE, cells = TRUE)
# colnames(nlcd_df) <- c("x", "y", "Land_Cover_Type")
```

```{r}
########################################
# NLCD Raster Processing and Conversion
########################################

# Reduce raster resolution
nlcd_ca_resampled <- aggregate(nlcd_ca, fact=10, fun=modal)  # Increase `fact` if necessary

# Convert to dataframe
nlcd_df <- as.data.frame(nlcd_ca_resampled, xy = TRUE, cells = TRUE)
colnames(nlcd_df) <- c("x", "y", "Land_Cover_Type")

# Check raster categories before conversion
print(unique(values(nlcd_ca_resampled)))

# Convert raster to polygon while preserving values
nlcd_poly <- as.polygons(nlcd_ca_resampled, dissolve = TRUE) %>%
  st_as_sf()

# Check column names
print(colnames(nlcd_poly))

# Rename the raster value column (which currently has the raster filename)
colnames(nlcd_poly)[colnames(nlcd_poly) == "Annual_NLCD_LndCov_2020_CU_C1V0"] <- "Land_Cover_Type"

# Ensure column exists
if (!"Land_Cover_Type" %in% colnames(nlcd_poly)) {
  stop("Land_Cover_Type column is missing in nlcd_poly. Check raster attributes.")
}
```

```{r}
# Plot using vectorized raster
ggplot() +
  geom_sf(data = nlcd_poly, aes(fill = factor(Land_Cover_Type)), color = NA) +
  geom_sf(data = st_transform(fire_perimeters, crs(nlcd_ca)), fill = NA, color = "black", linewidth = 0.3) +
  labs(title = "NLCD Land Cover and 2020 Fire Perimeters in California",
       fill = "Land Cover Type") +
  theme_minimal()
```

```{r}

# Define land cover type labels correctly
land_cover_labels <- data.frame(
  Land_Cover_Type = c(11, 12, 21, 22, 23, 24, 31, 41, 42, 43, 52, 71, 81, 82, 90, 95),
  Land_Cover_Name = c("Open Water", "Perennial Ice/Snow", "Developed, Open Space",
                      "Developed, Low Intensity", "Developed, Medium Intensity",
                      "Developed, High Intensity", "Barren Land",
                      "Deciduous Forest", "Evergreen Forest", "Mixed Forest",
                      "Shrub/Scrub", "Grassland/Herbaceous", "Pasture/Hay",
                      "Cultivated Crops", "Woody Wetlands",
                      "Emergent Herbaceous Wetlands")
)
```

```{r}
# # Ensure raster values are preserved before converting to polygons
# nlcd_poly <- as.polygons(nlcd_ca_resampled, dissolve = TRUE) %>%
#   st_as_sf()
# 
# # Rename raster column to match previous use
# colnames(nlcd_poly)[colnames(nlcd_poly) == "Annual_NLCD_LndCov_2020_CU_C1V0"] <- "Land_Cover_Type"

# Convert Land_Cover_Type to numeric for proper merging
nlcd_poly$Land_Cover_Type <- as.numeric(nlcd_poly$Land_Cover_Type)

# Merge with land cover labels
nlcd_poly <- left_join(nlcd_poly, land_cover_labels, by = "Land_Cover_Type")

##
#Check here if label naming works
nlcd_poly
##

# Categorize land use for Developed vs. Natural/Other
nlcd_poly <- nlcd_poly %>%
  mutate(
    Land_Use_Category = ifelse(Land_Cover_Type %in% c(21,22,23,24), 
                               "Developed (Urban/Populated)",
                               "Natural / Other")
  )

# Check if all labels joined correctly
if (!"Land_Cover_Name" %in% colnames(nlcd_poly)) {
  stop("Land_Cover_Name column is missing after merging labels.")
}
```

```{r}
# Plot using vectorized raster
ggplot() +
  geom_sf(data = nlcd_poly, aes(fill = factor(Land_Cover_Name)), color = NA) +
  geom_sf(data = st_transform(fire_perimeters, crs(nlcd_ca)), fill = NA, color = "black", linewidth = 0.3) +
  labs(title = "NLCD Land Cover and 2020 Fire Perimeters in California",
       fill = "Land Cover Type") +
  theme_minimal()
```

```{r}

# Plot with categorized land usage
ggplot() +
  geom_sf(data = nlcd_poly, aes(fill = Land_Use_Category), color = NA) +
  geom_sf(data = fire_perimeters, fill = NA, color = "red", linewidth = 0.4) +
  scale_fill_manual(values=c("gray", "forestgreen"), name="Land Usage Type") +
  labs(title = "California Land Cover (2020) and Fire Perimeters",
       subtitle = "Developed vs Natural areas affected by 2020 fires",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```

# Findings

Based on the code implementation, this analysis provided:

-   A ranked list of California cities most vulnerable to wildfire impacts based on proximity measurements

-   Visual representations of the spatial relationships between population centers and fire perimeters

-   Population weighted fire risk analysis

-   Quantitative breakdown of land cover types within fire-affected areas

-   Insights into whether fires predominantly affected natural landscapes or encroached on developed areas
