---
title: "Wildfires in California in 2020"
output: 
  html_document:
    toc: true          
    toc_depth: 4       
    number_sections: true 
date: '`r Sys.Date()`'
author: "Irache Vergara, Gretchen Moulton, Viktoria Gagua, Daniela Guggenberger"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Portion for Introduction


```{r}
rm(list = ls())
library(sf)
library(dplyr)
library(ggplot2)
library(raster)
library(lubridate)
library(terra)
library(vroom)
library(patchwork)
```

We will first load our data containing perimeters of all wildfires in California. We will then filter to just look at data from 2020. We will also load data containing the boundary of California. 
```{r}
# We will load our data containing spatial data for wildfires in 2020
url <- "/vsicurl/https://raw.githubusercontent.com/Viktoriag27/Geospatial_Cali_Wildfires/main/merged_fires_cleaned.gpkg"
fire_data <- st_read(url)

# We need to get the California Border shape file by getting US States
url <- "/vsicurl/https://github.com/Viktoriag27/Geospatial_Cali_Wildfires/raw/main/CA_Outline/ne_110m_admin_1_states_provinces/ne_110m_admin_1_states_provinces.shp"
US_States <- st_read(url)
# too many columns to work with, we will select fewer
US_States_simple <- US_States %>% dplyr::select(name, postal, latitude, longitude, geometry)
# select just CA 
california_boundary <- US_States_simple %>% filter(postal=="CA")

# Convert fire data to match California CRS
fire_data <- st_transform(fire_data, crs = st_crs(california_boundary)) 
fire_data <- st_make_valid(fire_data)
#simplify using 10 meters to make computations a bit quicker
fire_data<- st_simplify(fire_data, dTolerance = 10) 

#Let's see what this looks like
ggplot()+
 geom_sf(data=california_boundary, fill="lightblue", color = "black")+
 geom_sf(data=fire_data, color = "black") 
```

## Important Variables

To study the role of various factors on the intensity of wildfires, we will define the intensity of wildfires in a variety of ways: duration, size, and brightness. 

### Fire Duration (Days)

ADD WRITTEN DESCRIPTION OF VARIABLE HERE!

```{r}
#fill scale is number of days each fire lasted
ggplot()+
  geom_sf(data=california_boundary, fill="lightblue", color = "black")+
  geom_sf(data=fire_data, aes(fill = duration), color = "black") +
  scale_fill_gradient(low = "pink", high = "darkblue")  

```

The shortest fires lasted only 1 day and the longest fire lasted for 183 days. We can see the variation plotted above. 

There are 11 fires that have NAs for `CONT_DATE`, making their duration NA. We remove these observations, as it is an instance of missing data. Our data contains other observations of fires contained outside of 2020 (in 2021) so these NAs are not an issue of containment occurring outside of the calendar year. 

```{r}
# Count how many rows have a missing duration 
missing <- fire_data[is.na(fire_data$duration), ]
count(missing)
# Remove rows where duration is NA
fire_data <- fire_data[!is.na(fire_data$duration), ]
```

### Fire Brightness (Kelvins)

ADD WRITTEN DESCRIPTION OF VARIABLE HERE!

```{r}
#fill using the average brightness of fires
ggplot()+
  geom_sf(data=california_boundary, fill="lightblue", color = "black")+
  geom_sf(data=fire_data, aes(fill = avg_brightness), color = "black") +
  scale_fill_gradient(low = "pink", high = "yellow")  # Adjust gradient colors

```

Here we can see there is very little variation in the average brightness of each fire. The range of brightness of the fires is 301 K to 401 K. 

Many of our fires do not contain brightness data (384). We will store a new dataframe `fire_data_brightness` with only the observations that contain brightness (110). 

```{r}
#count how many missing 
missing <- fire_data[is.na(fire_data$avg_brightness), ]

count(missing)

#there are 384 observations that do not contain brightness data. We will store a new data frame with just the observations containing brightness 

fire_data_brightness <- fire_data %>%
  filter(!is.na(avg_brightness))
count(fire_data_brightness)
```

### Fire Size (Acres)

ADD WRITTEN DESCRIPTION OF VARIABLE HERE!

```{r}
#fill by size of fire
ggplot()+
  geom_sf(data=california_boundary, fill="lightblue", color = "black")+
  geom_sf(data=fire_data, aes(fill = GIS_ACRES), color = "black") +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen")  
```
Each fire contains data of the size of the fire in acres. There is no missing data for this variable. The fires in this data set ranged from 1 acre to 1,032,700 acres. 

## Weather: Temperature 
### Data Preparation and Exploration

We will first begin our investigation focusing on the role of weather, specifically air temperature on the intensity of fires. 

The US National Oceanic and Atmosphere Administration (NOAA) (https://www.noaa.gov/noaa-wildfire/wildfire-climate-connection) explains that climate change has lead to an increase in wildfires. More carbon emissions have led to an increase in weather shocks, such as droughts, and an overall warmer climate, creating hotter and drier conditions that are more prone to fire. To evaluate the impact of a warming planet on wildfires, we would need to evaluate year over year trends. However, we can also investigate the role of short term variations in the weather and to see their impact on wildfires. 

One way to do this is to look at fluctuation in temperature throughout one year. We gathered GeoTIFF files detailing daily maximum temperatures (tmax) measured in Celsius across the United States from NOAA. The data can be accessed here: https://ftp.cpc.ncep.noaa.gov/GIS/GRADS_GIS/GeoTIFF/TEMP/us_tmax/. Rather than look at raster data from 365 different days in 2020, which computationally was challenging due to the size of the GeoTIFF images, we chose to look at the 15th of each month to identify annual cyclical trends in temperature. We chose to work with this data as it was the most detailed and showed variation throughout California. When we looked at monthly averages, the national level data was not detailed enough and showed very little variation at the state level on average. We opted to prioritize detailed data to better visualize on a map the association between temperature and wildfire occurrences.

We recognize this approach has significant limitations. An ideal analysis of shocks in temperature would gather data for each day of 2020 to see better see extremes. The best way to study trends in temperature would be to look at monthly averages across decades. However, due to data limitations neither of these approaches were feasible. We note that in the future we could enhance the project by looking at daily or monthly average or max temperatures. 

To begin our analysis, we first need to identify which fires occurred in which months. To organize the data we will create 12 binary variables, 1 for each month, that designate if the fire occurred during that month.

```{r}
#create binary variables for each month containing 1 if a fire occurred in that month, 0 if not
fire_data <- fire_data %>%
  mutate(
    Jan = ifelse(between(month(ALARM_DATE), 1, 1) | between(month(CONT_DATE), 1, 1), 1, 0),
    Feb = ifelse(between(month(ALARM_DATE), 2, 2) | between(month(CONT_DATE), 2, 2), 1, 0),
    Mar = ifelse(between(month(ALARM_DATE), 3, 3) | between(month(CONT_DATE), 3, 3), 1, 0),
    Apr = ifelse(between(month(ALARM_DATE), 4, 4) | between(month(CONT_DATE), 4, 4), 1, 0),
    May = ifelse(between(month(ALARM_DATE), 5, 5) | between(month(CONT_DATE), 5, 5), 1, 0),
    Jun = ifelse(between(month(ALARM_DATE), 6, 6) | between(month(CONT_DATE), 6, 6), 1, 0),
    Jul = ifelse(between(month(ALARM_DATE), 7, 7) | between(month(CONT_DATE), 7, 7), 1, 0),
    Aug = ifelse(between(month(ALARM_DATE), 8, 8) | between(month(CONT_DATE), 8, 8), 1, 0),
    Sep = ifelse(between(month(ALARM_DATE), 9, 9) | between(month(CONT_DATE), 9, 9), 1, 0),
    Oct = ifelse(between(month(ALARM_DATE), 10, 10) | between(month(CONT_DATE), 10, 10), 1, 0),
    Nov = ifelse(between(month(ALARM_DATE), 11, 11) | between(month(CONT_DATE), 11, 11), 1, 0),
    Dec = ifelse(between(month(ALARM_DATE), 12, 12) | between(month(CONT_DATE), 12, 12), 1, 0)
  )
# sum to see how many fires occurred in each month
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
monthly_sums <- sapply(months, function(month) sum(fire_data[[month]], na.rm = TRUE))
monthly_sums

```

From this we can see most fires occur in the summer and fall (May - November), with August being the month with the most fires in 2020. 

Now we will explore to see the monthly temperatures and the fires that occurred in that month.  We first load our GeoTiffs for each month. We plot the weather raster for the 15th of each month and the fires that occurred in each month. 

```{r}
# Define the base URL for the raster files in GitHub
github_base <- "/vsicurl/https://github.com/Viktoriag27/Geospatial_Cali_Wildfires/raw/main/Weather/Temperature/"

# this is our list of months
months <- c("January", "February", "March", "April", "May", "June", 
            "July", "August", "September", "October", "November", "December")

# Create a named list with the updated GitHub paths
monthly_rasters <- setNames(
  paste0(github_base, months, "_2020.tif"),
  substr(months, 1, 3) 
)
# gather all raster files into a list
raster_list <- lapply(monthly_rasters, rast)

# transform California shapefile to match the CRS of the rasters
california_boundary <- st_transform(california_boundary, crs = 4326)

# process rasters and crop to California shape
processed_rasters <- lapply(monthly_rasters, function(raster_path) {
  temp_raster <- rast(raster_path)
  template_raster <- rast(temp_raster)
  res(template_raster) <- 0.02
  ext(template_raster) <- ext(temp_raster)
  temp_raster_resampled <- resample(temp_raster, template_raster, method = "cubic")
  temp_raster_cropped <- crop(temp_raster_resampled, ext(california_boundary))
  california_boundary_r <- rasterize(vect(california_boundary), temp_raster_cropped, field = 1)
  mask(temp_raster_cropped, california_boundary_r)
})

# we also want to group months by season for simpler plotting
seasons <- list(
  Winter = c("Dec", "Jan", "Feb"),
  Spring = c("Mar", "Apr", "May"),
  Summer = c("Jun", "Jul", "Aug"),
  Fall = c("Sep", "Oct", "Nov")
)

# plot each season
for (season in names(seasons)) {
  par(mfrow = c(1, 3), mar = c(5, 5, 5, 1))
  for (month in seasons[[season]]) {
    temp_raster_precise <- processed_rasters[[month]]
    plot(temp_raster_precise, main = paste(month, "2020 Max Temperature in California"),
         col = colorRampPalette(c("blue", "cyan", "yellow", "red"))(100), cex.main = 0.8)
    plot(st_geometry(california_boundary), add = TRUE, border = "black", lwd = 2)
    
    # Overlay fire data 
    if (month %in% names(fire_data) && any(fire_data[[month]] == 1, na.rm = TRUE)) {
      fires_this_month <- fire_data %>% filter(fire_data[[month]] == 1)
      if (nrow(fires_this_month) > 0) {
        fire_sf <- st_as_sf(fires_this_month, coords = c("Longitude", "Latitude"), crs = 4326)
        plot(st_geometry(fire_sf), add = TRUE, border = "black", col = NA, cex = 1.5, lwd = 1.3)
      }
    }
  }
  mtext(season, side = 3, line = -2, outer = TRUE, cex = 1)
}

```

Here we see the temperature across California on the 15th of the month and the fires that occurred in each month. From these graphs, we can easily visualize the seasonality of the fires, with more wildfires occurring in summer and fall. 

Now let's specifically investigate the average max temperature of the area of the fire. We will use extraction of the data from the rasters to conduct this step. We will use temperature data from the area of the fire from the months the fire occurred. For example, if the fire occurred only in July, we will find the average max temperature across the area where it occurred on July 15th. If the fire occurred in July and August, we will find the average max temperature in the area of the fire from July 15th and August 15th. Again, we note a limitation of this data is the fire may not have occurred on the 15th of the month. Max or average daily temperatures would better encapsulate actual temperatures when the fire occurred. 

```{r}
# Initialize a column for average temperature in the fire dataset
fire_data$avg_temp <- NA

# loop through each month
for (month in names(processed_rasters)) {
  temp_raster <- processed_rasters[[month]]
  # Ensure fire data CRS matches raster CRS
  fire_data <- st_transform(fire_data, crs = crs(temp_raster))
  # Filter on fires that occurred in the month
  valid_indices <- which(!is.na(fire_data[[month]]) & fire_data[[month]] == 1)
  if (length(valid_indices) > 0) {
    fires_this_month <- fire_data[valid_indices, ]
    # Convert fire polygons to terra format (SpatVector)
    fire_vect <- vect(fires_this_month)
    # Extract temperature values within each fire polygon and compute the mean
    extracted_values <- terra::extract(temp_raster, fire_vect, fun = function(x) { if (all(is.na(x))) NA else mean(x, na.rm = TRUE) })
    # Check if extracted values exist
    if (!is.null(extracted_values)) {
      # Assign the extracted temperature values back to the fire dataset
      fire_data$avg_temp[valid_indices] <- extracted_values[, 2]
    } else {
      print(paste("Warning: No temperature data extracted for fires in", month))
    }
  } else {
    print(paste("No fires occurred in", month))
  }
}
# There are 8 small fires that occur outside the border of California so we drop these
fire_data <- fire_data %>%
  filter(!is.na(avg_temp))

# Plot California map with fire polygons
ggplot() +
  geom_sf(data = california_boundary, fill = "lightgrey", color = "black") +  
  geom_sf(data = fire_data, aes(fill = avg_temp), color = "black", alpha = 0.7) + 
  scale_fill_gradientn(colors = colorRampPalette(c("blue", "cyan", "yellow", "red"))(100), name = "Avg Max Temp (°C)" ) +  
  labs(title = "Average Max Temperature During Wildfires (2020)",
       subtitle = "Based on Max Temperature Data for the 15th of the Month") +
  theme_minimal() +
  theme(legend.position = "right")
```

Here we have plotted all of the fires from 2020 filled with the average temperature on the 15th of each month they occurred during to see the differences in temperature for different fires. From here we can see that some of the larger fires do not appear be occurring during warmer temperatures, with the biggest fire having an average max temperature (outside air) of around 15 degrees. 

Now we will use the average max temperature in the area of each fire we just calculated to investigate how it may influence fire intensity: the size of the fire (acreage), duration of fire (days), and the brightness of each fire (Kelvin (K).

#### Temperature and Fire Size

We will investigate the role of temperature on the size of fires. We are going to look at each season individually because the data is incredibly skewed to the right as most fires occur in the summer and fall when it is warmer. Therefore it is challenging to assess any trend when looking at the data all together. 

```{r}
#We filter to get fires occurring in each season
summer_fires <- fire_data %>%
  filter(Jun == 1 | Jul == 1 | Aug == 1)
fall_fires <- fire_data %>%
  filter(Sep == 1 | Oct == 1 | Nov == 1)
winter_fires <- fire_data %>%
  filter(Dec == 1 | Jan == 1 | Feb == 1)
spring_fires <- fire_data %>%
  filter(Mar == 1 | Apr == 1 | May == 1)

# Scatter plot for Summer Fires
plot_summer <- ggplot(summer_fires, aes(x = avg_temp, y = GIS_ACRES)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Summer",
       x = "Avg Temperature (°C)", y = "Fire Size (Acres)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Fall Fires
plot_fall <- ggplot(fall_fires, aes(x = avg_temp, y = GIS_ACRES)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Fall",
       x = "Avg Temperature (°C)", y = "Fire Size (Acres)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Winter Fires
plot_winter <- ggplot(winter_fires, aes(x = avg_temp, y = GIS_ACRES)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Winter",
       x = "Avg Temperature (°C)", y = "Fire Size (Acres)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Spring Fires
plot_spring <- ggplot(spring_fires, aes(x = avg_temp, y = GIS_ACRES)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Spring",
       x = "Avg Temperature (°C)", y = "Fire Size (Acres)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Load patchwork library
library(patchwork)

# Combine all the plots in one row using patchwork with spacing adjustments
(plot_summer + plot_fall + plot_winter + plot_spring) + 
  plot_layout(ncol = 2) + 
  plot_layout(nrow = 2) +
  plot_annotation(title = "Average Temperature vs. Fire Size by Season") &
  theme(plot.title = element_text(hjust = 0.5))

```

Here we see a negative trend across all seasons. As temperatures rise, it appears fire size decreases; however, this may be  largely driven by extreme outliers. We will now remove outliers from our data as most fires are less than 250,000 acres. We will also filter on fires that exceed 1,000 acres because any wildfire below 1,000 acres is classified as small. We will focus on these large and medium wildfires, as it is difficult to show trends when including extremely large or small fires. 
```{r}
# Filter the fire data to include only fires with GIS_ACRES < 300,000 acres
fire_data_filtered <- fire_data %>%
  filter(GIS_ACRES < 250000) %>%
  filter(GIS_ACRES > 1000)

summer_fires_filtered <- fire_data_filtered %>%
  filter(Jun == 1 | Jul == 1 | Aug == 1)
fall_fires_filtered <- fire_data_filtered %>%
  filter(Sep == 1 | Oct == 1 | Nov == 1)
winter_fires_filtered <- fire_data_filtered %>%
  filter(Dec == 1 | Jan == 1 | Feb == 1)
spring_fires_filtered <- fire_data_filtered %>%
  filter(Mar == 1 | Apr == 1 | May == 1)


# Scatter plot for Summer Fires
plot_summer <- ggplot(summer_fires_filtered, aes(x = avg_temp, y = GIS_ACRES)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Summer",
       x = "Avg Temperature (°C)", y = "Fire Size (Acres)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Fall Fires
plot_fall <- ggplot(fall_fires_filtered, aes(x = avg_temp, y = GIS_ACRES)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Fall",
       x = "Avg Temperature (°C)", y = "Fire Size (Acres)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Winter Fires
plot_winter <- ggplot(winter_fires_filtered, aes(x = avg_temp, y = GIS_ACRES)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Winter",
       x = "Avg Temperature (°C)", y = "Fire Size (Acres)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Load patchwork library
library(patchwork)

# Combine all the plots in one row using patchwork with spacing adjustments
(plot_summer + plot_fall + plot_winter) + 
  plot_layout(ncol = 3) + 
  plot_annotation(title = "Average Temperature vs. Fire Size by Season") &
  theme(plot.title = element_text(hjust = 0.5))

```

From this analysis, we notice a positive trend for wildfires in the winter, but there is a negative trend for summer and fall. There were no fires in spring that were larger than 1,000 acres so we did not include this in our analysis. It appears that during the warmer months (summer and fall), hotter temperatures are associated with smaller fires. It may be that fires are more likely to occur when its hot, so officials are more prepared to respond faster and thus the fires are put out quicker. Additional research needs to be conducted, ideally with daily temperature data or monthly averages to assess these trends. 

#### Temperature and Fire Duration

```{r}
# Scatter plot for Summer Fires
plot_summer <- ggplot(summer_fires, aes(x = avg_temp, y = duration)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Summer",
       x = "Avg Temperature (°C)", y = "Fire Duration (Days)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Fall Fires
plot_fall <- ggplot(fall_fires, aes(x = avg_temp, y = duration)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Fall",
       x = "Avg Temperature (°C)", y = "Fire Duration (Days)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Winter Fires
plot_winter <- ggplot(winter_fires, aes(x = avg_temp, y = duration)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Winter",
       x = "Avg Temperature (°C)", y = "Fire Duration (Days)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Spring Fires
plot_spring <- ggplot(spring_fires, aes(x = avg_temp, y = duration)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Spring",
       x = "Avg Temperature (°C)", y = "Fire Duration (Days)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Load patchwork library
library(patchwork)

# Combine all the plots in one row using patchwork with spacing adjustments
(plot_summer + plot_fall + plot_winter + plot_spring) + 
  plot_layout(ncol = 2) + 
  plot_layout(nrow = 2) +
  plot_annotation(title = "Average Temperature vs. Fire Duration by Season") &
  theme(plot.title = element_text(hjust = 0.5))



```

When first looking at the plots, we can see that as temperatures rise in the summer and fall, the duration of the fire decreases; however, this may be more indicative that during hotter times, the number of small fires increases. Nevertheless, in the summer and fall, we see a negative correlation. However in winter, we see a positive trend, where hotter temperatures are associated with longer fires. In spring there is no trend. 

It appears most fires are less than 50 days. We will investigate fires under 50 days to see if there are any other discernible trends once some outliers are removed. 

```{r}
# Filter the fire data to include only fires lasting less than 50 days
fire_data_under_50days <- fire_data %>%
  filter(duration < 50)

summer_fires_50days <- fire_data_under_50days %>%
  filter(Jun == 1 | Jul == 1 | Aug == 1)
fall_fires_50days <- fire_data_under_50days %>%
  filter(Sep == 1 | Oct == 1 | Nov == 1)
winter_fires_50days <- fire_data_under_50days %>%
  filter(Dec == 1 | Jan == 1 | Feb == 1)
spring_fires_50days <- fire_data_under_50days %>%
  filter(Mar == 1 | Apr == 1 | May == 1)


# Scatter plot for Summer Fires
plot_summer <- ggplot(summer_fires_50days, aes(x = avg_temp, y = duration)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Summer",
       x = "Avg Temperature (°C)", y = "Fire Duration (Days)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Fall Fires
plot_fall <- ggplot(fall_fires_50days, aes(x = avg_temp, y = duration)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Fall",
       x = "Avg Temperature (°C)", y = "Fire Duration (Days)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Winter Fires
plot_winter <- ggplot(winter_fires_50days, aes(x = avg_temp, y = duration)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Winter",
       x = "Avg Temperature (°C)", y = "Fire Duration (Days)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Spring Fires
plot_spring <- ggplot(spring_fires_50days, aes(x = avg_temp, y = duration)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Spring",
       x = "Avg Temperature (°C)", y = "Fire Duration (Days)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Load patchwork library
library(patchwork)

# Combine all the plots in one row using patchwork with spacing adjustments
(plot_summer + plot_fall + plot_winter + plot_spring) + 
  plot_layout(ncol = 2) + 
  plot_layout(nrow = 2) +
  plot_annotation(title = "Average Temperature vs. Fire Duration by Season") &
  theme(plot.title = element_text(hjust = 0.5))

```

Once we remove fires that have a duration over 50 days, we see that the plots for summer and fall change drastically, with fall having no correlation and summer still has a negative trend but less steep. Spring still has no discernible trend and the positive trend in winter also became drastically less steep. 

There are no clear trends from this analysis once we removed outliers. It appears that temperature may not play a key role in the duration of a fire in our data.  

#### Temperature and Fire Brightness

We will now analyze to see if there is any correlation between temperature and brightness of fires. 

```{r}
# Need to recreate fire_data_brightness so that it includes avg_temp variable
fire_data_brightness <- fire_data %>%
  filter(!is.na(avg_brightness))

# organize data for each season
summer_fires_brightness <- fire_data_brightness %>%
  filter(Jun == 1 | Jul == 1 | Aug == 1)
fall_fires_brightness <- fire_data_brightness %>%
  filter(Sep == 1 | Oct == 1 | Nov == 1)
winter_fires_brightness <- fire_data_brightness %>%
  filter(Dec == 1 | Jan == 1 | Feb == 1)
spring_fires_brightness <- fire_data_brightness %>%
  filter(Mar == 1 | Apr == 1 | May == 1)


# Scatter plot for Summer Fires
plot_summer <- ggplot(summer_fires_brightness, aes(x = avg_temp, y = avg_brightness)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Summer",
       x = "Avg Temperature (°C)", y = "Fire Brightness (Kelvin)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Fall Fires
plot_fall <- ggplot(fall_fires_brightness, aes(x = avg_temp, y = avg_brightness)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Fall",
       x = "Avg Temperature (°C)", y = "Fire Brightness (Kelvin)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Winter Fires
plot_winter <- ggplot(winter_fires_brightness, aes(x = avg_temp, y = avg_brightness)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Winter",
       x = "Avg Temperature (°C)", y = "Fire Brightness (Kelvin)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Scatter plot for Spring Fires
plot_spring <- ggplot(spring_fires_brightness, aes(x = avg_temp, y = avg_brightness)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Spring",
       x = "Avg Temperature (°C)", y = "Fire Brightness (Kelvin)") +
  theme_minimal()+
  theme(plot.title = element_text(size = 10),
  axis.title.x = element_text(size = 8), 
  axis.title.y = element_text(size = 8) )

# Load patchwork library
library(patchwork)

# Combine all the plots in one row using patchwork with spacing adjustments
(plot_summer + plot_fall + plot_winter + plot_spring) + 
  plot_layout(ncol = 2) + 
  plot_layout(nrow = 2) +
  plot_annotation(title = "Average Temperature vs. Fire Brightness by Season") &
  theme(plot.title = element_text(hjust = 0.5))

```

Here we see there be a very slight positive correlation between temperature and fire brightness in the summer; however, in the fall we see a slight negative correlation. There is very little data for winter and spring and therefore no trends are evident. From this we can conclude that temperature does not appear to play a large role on fire brightness in our data set; however, it should be noted that a main limitation in this analysis is that only 110 variables had data on brightness, as we had to merge this data from a different fire data set (see Appendix). Perhaps if more fires had brightness data, trends would be present in the data. 

#### Summary

From this analysis, it does not appear that higher temperatures lead to more intense fires. However, it is important to emphasize this may be due to limitations in the data. As previously discussed, the temperature data is not ideal for analyzing temperatures shocks. This data only reflects the 15th of each month and looking at daily max temperatures and highlighting extreme temperature shocks from the average would perhaps have different results.

- Fire Size and Temperature: A negative trend was initially observed across all seasons, where fire size seemed to decrease as temperatures rise. However, this relationship was likely influenced by extreme outliers. After removing outliers and focusing on fires exceeding 1,000 acres (to exclude small wildfires), we noticed a positive trend in fire size during the winter, but slight negative trends in the summer and fall. Further research may be able to better validate these trends. 

- Fire Duration and Temperature: For the summer and fall seasons, as temperatures increase, fire duration tends to decrease. However, in winter, there was a positive trend, and spring shows no clear relationship between temperature and fire duration. When we analyzed only fires lasting less than 50 days, no relationship was identified in any season. 

- Fire Brightness and Temperature: A very slight positive correlation was observed between temperature and fire brightness during the summer, while the fall showed a slight negative correlation. There was insufficient data in the winter and spring to discern any significant trends.

Overall, the relationship between temperature and wildfires shows seasonal variation, but we cannot establish a clear pattern between temperature and wildfire intensity. We will now explore another aspect of weather that may influence wildfire intensity. 

